Is high social class always beneficial for survival? A study of northern Sweden 1801--2013 \footnote{Short title, "Is high social class beneficial for survival?"}

### 2018-10-09 popstud_anon.Rmd
copied from 'open' and modified

---
title: "Is high social class always beneficial for survival? A study of northern Sweden 1801--2013"
author: X \footnote{Corresponding author; Phone number +0000000000, email X@xcom.com}
  and Y
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: yes
    toc_depth: 2
citation_package: natbib
classoption: titlepage
biblio-style: apalike
documentclass: article
fontsize: 12pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{endfloat}
- \usepackage[doublespacing]{setspace}
- \usepackage{sectsty}
- \subsectionfont{\normalfont\itshape}
bibliography: mybib
abstract: "Focusing on two regions in northern Sweden 1801--2013, we challenge common
  notions of the assumed advantage in survival of belonging to a high social class.
  The issue is analysed according to gender and age group (adults and elderly) and
  in relation to the development of economic inequality. The results show that high
  social class is not always favourable for survival. Men in the elite category, particularly
  in working age, had higher mortality compared to others during a large part of the
  studied period; a male mortality class reversal appears at a surprisingly late date,
  while the social gradient among women conforms to the expected pattern. We suggest
  that health-related behaviour is decisive not only in later but earlier phases of
  the mortality transition as well. The results implicate that the association between
  social class and health is more complex than is assumed in many of the dominant
    theories in demography and epidemiology.\\par\n\\textbf{Keywords:} Class reversal
  in mortality; Cox regression; Gender differences; Inverse probability weighting;
  Mortality transition; Northern Sweden; Social inequality; 19th to 21 century."

titlepage: yes
---

@@

@@

The data for the present study come from two large population databases at the Demographic Data Base (DDB),Umeå University (http://www.cedar.umu.se), which provide us with micro-data for the Skellefteå and Umeå regions in northern Sweden. Our data set constitutes a large population with substantial social diversity that can be followed for a uniquely long time. The period 1801--1950 is covered by the database Poplink [@wee16]. Poplink is based on linked parish records, allowing us to reconstruct life biographies on people as long as they remained in the region. The records are linked within, but not
between, the regions.

@@

Figures \@ref(fig:hiscskumm) and \@ref(fig:hiscskumw) show the distribution
(per cent) of exposure time according to class for men and women,
respectively. The farmer category was dominant until the middle of the 20th centuries but have become marginal in the latest periods, particularly among women. The category of workers is fairly stable over time. The highest class has increased substantially, albeit from a very low level
in the 19th century. The increase in academics in connection to the
establishment of Umeå University explains much of this increase in later time. Furthermore,
we observe a rather high percentage of missing social class in the early 19th century.
For men the proportion is somewhat higher in the age group 65 and above, those who were retired,
but still low enough to be used in our analyses (results not shown). If we disregard the very first period,
however, there were very few with occupation missing until 1950.

```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, none) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = "14cm", fig.height=4}
xx <- with(vb[vb$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xy <- with(skume[skume$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xz <- with(foli[foli$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xxx <- rbind(xx, xy, xz)
yy <- round(prop.table(xx, 1) * 100, 1)
barplot(t(yy)[5:1, ], col = c("red", "darkgreen", "blue", "green", "black"), las = 2, ylab = "Per cent")
```

```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, none) and time period, women.", out.height = "14cm", fig.scap="Exposure by social class and time period, women.",fig.height=4}
xx <- with(vb[vb$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xy <- with(skume[skume$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xz <- with(foli[foli$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xxx <- rbind(xx, xy, xz)
yy <- round(prop.table(xx, 1) * 100, 1)
barplot(t(yy)[5:1, ], col = c("red", "darkgreen", "blue", "green", "black"), las = 2, ylab = "Per cent")
```

The proportion with missing values among women is a larger problem, however;
see Figure \@ref(fig:hiscskumw). Corresponding to what we found for men, the
proportion without social position is very low until the middle of the 20th
century, reflecting the way women's social position is defined. However, the frequency of missing social class is quite high during particularly the first two periods of data from the Linnaeus database---from 1976. Since female social class for this period is defined
according to own occupation, the consequence of many women being outside
the workforce is that the proportion of missing increases. For the later periods,
the results are fairly complete.

@@

From Table \@ref(tab:cumswtab) it is obvious that the *statistical significance* is strong in our statistical model of the probability of dying. The most interesting feature is the three-way interaction with **sex**, **period**, and **social class**. This indicates that fundamental changes have taken place in the relationship between the variables during our study period.

```{r probswabs, fig.cap = "Absolute risk of dying before the age of 65, given survival to age 40, by period and social class.", fig.scap = "Absolute risk of dying before 65.", fig.height = 8, out.height = "14cm"}
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
refclw <- 1 ## pdw[3, ]
plot(midper, pdw[1, ] / refclw, col = scol[1], type = "b", ylim = c(0.0, 0.55),
     xlab = "Period", xlim = c(1800, 2015), ##log = "y",
     ylab = "P(die before 65 | alive at 40", axes = FALSE, main = "Women, 40-64", lty = 1, pch = 1)

axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
axis(2, las = 1)
axis(4, las = 1)
abline(h = 0)
box()
for (i in 2:5){
    lines(midper, pdw[i, ] / refclw, col = scol[i], type = "b", pch = i, lty = i)
}
legend("bottomleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
refclm <- 1 ##pdm[3, ]
plot(midper, pdm[1, ] / refclm, col = scol[1], type = "b", ylim = c(0.0, 0.55), xlab = "Period",
     ylab = "P(die before 65 | alive at 40)", axes = FALSE, main = "Men, 40-64", lty = 1, pch = 1,
     xlim = c(1800, 2015))#, log = "y")
axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
axis(2, las = 1)
axis(4, las = 1)
abline(h = 0)
box()
for (i in 2:5){
    lines(midper, pdm[i, ] / refclm, col = scol[i], type = "b", lty = i, pch = i)
}
legend("bottomleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
  ##              main = "Women 30-64", ylim = c(0, 1))##, exclude = 5:6)
##moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|30)",
  ##              main = "Men 30-64", ylim = c(0, 1))##, exclude = 5:6)
par(oldpar)
```

```{r probsw, fig.cap = "Relative risk of dying before the age of 65, given survival to age 40, by period and social class. Reference is worker.", fig.scap = "Relative risk of dying before 65.", fig.height = 8, out.height = "14cm"}
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
refclw <- pdw[3, ]
plot(midper, pdw[1, ] / refclw, col = scol[1], type = "b", ylim = c(1/3, 3),
     xlab = "Period", xlim = c(1800, 2015), log = "y",
     ylab = "Relative risk", axes = FALSE, main = "Women, 40-64", lty = 1, pch = 1)

axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
axis(2, las = 1)
axis(4, las = 1)
abline(h = 0)
box()
for (i in 2:5){
    lines(midper, pdw[i, ] / refclw, col = scol[i], type = "b", pch = i, lty = i)
}
legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
refclm <- pdm[3, ]
plot(midper, pdm[1, ] / refclm, col = scol[1], type = "b", ylim = c(1/3, 3), xlab = "Period",
     ylab = "Relative risk", axes = FALSE, main = "Men, 40-64", lty = 1, pch = 1,
     xlim = c(1800, 2015), log = "y")
axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.5)
axis(2, las = 1)
axis(4, las = 1)
abline(h = 0)
box()
for (i in 2:5){
    lines(midper, pdm[i, ] / refclm, col = scol[i], type = "b", lty = i, pch = i)
}
legend("topleft", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 0.7, bty = "n")
##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
  ##              main = "Women 30-64", ylim = c(0, 1))##, exclude = 5:6)
##moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|30)",
  ##              main = "Men 30-64", ylim = c(0, 1))##, exclude = 5:6)
par(oldpar)
```

```{r cumswtab, message=FALSE, cache = TRUE, eval = TRUE}
fitt <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex + civst + urban, data = vb1)

# vb3064$social_class <- vb3064$hisclass
# vb3064$marital_status <- vb3064$civst
# vb3064$urban_or_rural <- vb3064$urban
# vb3064$age_group <- vb3064$age
# fitt <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
#             data = vb3064, family = poisson)
dr <- drop1(fitt, test = "Chisq")
# ##dr
knitr::kable(dr, caption = "Single term deletions, ages 30-64.", digits = c(0, 1, 2, 4))
# ##print(xtable(dr,label = "tab:cumswtab", caption = "Mortality: ANOVA, ages 30--64."),
  ##    type = "latex", comment = FALSE)
```

@@

We now turn to the elderly, those who in most cases have left the workforce.
We use the *expected remaining years of life at 65 up to age 100* (reported as mean age at death) as the summary measure, since the relative risk is a not so good measure when risks are high. First, looking at Table \@ref(tab:cumswtabold), it is obvious that the statistical significance regarding differences between groups over time is strong.

```{r cumswtabold, cap="Table", cache = TRUE, eval = TRUE}
# vb6599$marital_status <- vb6599$civst
# vb6599$social_class <- vb6599$hisclass
# vb6599$urban_or_rural <- vb6599$urban
# vb6599$age_group <- vb6599$age
# fittold <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
#             data = vb6599, family = poisson)

fittold <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex + civst + urban, data = vb2)
drold <- drop1(fittold, test = "Chisq")
knitr::kable(drold, caption = "Single term deletions, ages 65-99.", digits = c(0, 1, 2, 4))
##drold
##print(xtable(drold,label = "tab:cumswtabold", caption = "Mortality: ANOVA, ages 65--99."),
  ##    type = "latex", comment = FALSE)
```

@@

This work was undertaken within the research programme *Ageing and
Living Conditions* at the Centre for Demographic and Ageing Research
(CEDAR) at Umeå University, financed by the Swedish Research Council (grant
No 349-2008-6592) and Edvinsson was economically supported by the Swedish Research Council via the project *Old age, health and inequality:
Mortality from a life course perspective in Sweden 1986--2009* (grant No 421-2012-892). We thank the
Demographic Data Base, Umeå University, for giving us access to data and
Jonas Helgertz at the Department of Economic History, Lund University, for
providing us with scripts for creating Hisclass from Swedish census
data.
-->

# References

>>>>

---
title: Is high social class always beneficial for survival? A study of northern Sweden 1801--2013 \footnote{Short title, "Is high social class beneficial for survival?"} 
author: X (corresponding author) and Y
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
  bookdown::word_document2:
    toc: yes
    toc_depth: 2
citation_package: natbib
classoption: titlepage
biblio-style: apalike
documentclass: article
fontsize: 12pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{endfloat}
- \usepackage[doublespacing]{setspace}
- \usepackage{sectsty}
- \subsectionfont{\normalfont\itshape}
bibliography: mybib
titlepage: yes
---

@@

\begin{abstract} 
Focusing on two regions in northern Sweden 1801--2013, we challenge common
  notions of the assumed advantage in survival of belonging to a high social class.
  The issue is analysed according to gender and age group (adults and elderly) and
  in relation to the development of economic inequality. The results show that high
  social class is not always favourable for survival. Men in the elite category, particularly
  in working age, had higher mortality compared to others during a large part of the
  studied period; a male mortality class reversal appears at a surprisingly late date,
  while the social gradient among women conforms to the expected pattern. We suggest
  that health-related behaviour is decisive not only in later but earlier phases of
  the mortality transition as well. The results implicate that the association between
  social class and health is more complex than is assumed in many of the dominant
    theories in demography and epidemiology.

    \textbf{Keywords:} Class reversal
  in mortality; Cox regression; Gender differences; Inverse probability weighting;
  Mortality transition; Northern Sweden; Social inequality; 19th to 21th century.

\end{abstract}

@@

The data for the present study come from two large population databases at the Demographic Data Base (DDB),Umeå University (http://www.cedar.umu.se), which provide us with micro-data for the Skellefteå and Umeå regions in northern Sweden [@ddb15]. The period 1801--1950 is covered by the database Poplink [@wee16]. Poplink is based on linked parish records, allowing us to reconstruct life biographies on people as long as they remained in the region. The records are linked within, but not
between, the regions.

@@

Figure \@ref(fig:plottaalla) shows the distribution
(per cent) of exposure times according to class for women and men (see also Tables \@ref(tab:womenexp) and \@ref(tab:menexp), respectively, as well as Table \@ref(tab:womendeaths), and \@ref(tab:mendeaths) for number of deaths during the different periods in the Appendix). The farmer category was dominant until the middle of the 20th centuries but have become marginal in the latest periods, particularly among women. The category of workers is fairly stable over time. The highest class has increased substantially, albeit from a very low level
in the 19th century. The increase in academics in connection to the
establishment of Umeå University explains much of this increase in later time. Furthermore,
we observe a rather high percentage of missing social class in the early 19th century.
For men the proportion is somewhat higher in the age group 65 and above, those who were retired,
but still low enough to be used in our analyses (results not shown). If we disregard the very first period,
however, there were very few with occupation missing until 1950.

<!--
```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, none) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = "14cm", fig.height=4,cache=FALSE}
cls <- rainbow(5, s = 1.0)
cls <- cls[c(2, 3, 1, 5, 4)]
cls[1] <- "black"
oldcols <- c("red", "darkgreen", "blue", "green", "black")
xx <- with(vb[vb$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xy <- with(skume[skume$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xz <- with(foli[foli$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xxx <- rbind(xx, xy, xz)
yy <- round(prop.table(xx, 1) * 100, 1)
##barplot(t(yy)[5:1, ], col = cls, las = 2, ylab = "Per cent")
```

```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, none) and time period, women.", out.height = "14cm", fig.scap="Exposure by social class and time period, women.",fig.height=4,cache=FALSE}
xx2 <- with(vb[vb$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xy <- with(skume[skume$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xz <- with(foli[foli$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
##xxx <- rbind(xx, xy, xz)
yy2 <- round(prop.table(xx2, 1) * 100, 1)
##barplot(t(yy2)[5:1, ], col = cls, las = 2, ylab = "Per cent")
```
-->

```{r plottaalla,fig.height = 8, out.height = "18cm", fig.cap="Relative exposure by social class (from the top: elite, middle, workers, farmers, none), time period, and sex.", fig.scap="Exposure by social class, time period, and sex"}
cls <- rainbow(5, s = 1.0)
cls <- cls[c(2, 3, 1, 5, 4)]
cls[1] <- "black"

oldpar <- par(mfrow = c(2, 1))
barplot(t(yy2)[5:1, ], col = cls, las = 2, ylab = "Per cent", main = "Women")
barplot(t(yy)[5:1, ], col = cls, las = 2, ylab = "Per cent", main = "Men")
par(oldpar)
```

The proportion of missing values among women is a larger problem, however. Corresponding to what we found for men, the
proportion without social position is very low until the middle of the 20th
century, reflecting the way women's social position is defined. However, the frequency of missing social class is quite high during particularly the first two periods of data from the Linnaeus database---from 1976. Since female social class for this period is defined
according to own occupation, the consequence of many women being outside
the workforce is that the proportion of missing increases. For the later periods,
the results are fairly complete.

@@

From Table \@ref(tab:cumswtabout) it is obvious that the *statistical significance* is strong in our statistical model of the probability of dying. The most interesting feature is the three-way interaction with **sex**, **period**, and **social class**. This indicates that fundamental changes have taken place in the relationship between the variables during our study period.

```{r probswabs, fig.cap = "Probability of dying before the age of 65, given survival to age 40, by period and social class.", fig.scap = "Probability of dying before 65.", fig.height = 10, out.height = "18cm", cache = FALSE}
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1), lwd = 1.5)
midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
refclw <- 1 ## pdw[3, ]
plot(midper, pdw[1, ] / refclw, col = scol[1], type = "b", ylim = c(0.0, 0.55),
     xlab = "", xlim = c(1800, 2015), ##log = "y",
     ylab = "P(die before 65 | alive at 40", axes = FALSE, main = "Women, 40-64", lty = 1, pch = 1)

axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.8)
axis(2, las = 1)
axis(4, las = 1)
abline(h = 0)
box()
for (i in 2:5){
    lines(midper, pdw[i, ] / refclw, col = scol[i], type = "b", pch = i, lty = i)
}
legend("topright", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 1, bty = "n")
refclm <- 1 ##pdm[3, ]
plot(midper, pdm[1, ] / refclm, col = scol[1], type = "b", ylim = c(0.0, 0.55), xlab = "",
     ylab = "P(die before 65 | alive at 40)", axes = FALSE, main = "Men, 40-64", lty = 1, pch = 1,
     xlim = c(1800, 2015))#, log = "y")
axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.8)
axis(2, las = 1)
axis(4, las = 1)
abline(h = 0)
box()
for (i in 2:5){
    lines(midper, pdm[i, ] / refclm, col = scol[i], type = "b", lty = i, pch = i)
}
legend("topright", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 1, bty = "n")
##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
  ##              main = "Women 30-64", ylim = c(0, 1))##, exclude = 5:6)
##moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|30)",
  ##              main = "Men 30-64", ylim = c(0, 1))##, exclude = 5:6)
par(oldpar)
```

```{r probsw, fig.cap = "Relative risk of dying before the age of 65, given survival to age 40, by period and social class. Reference is worker.", fig.scap = "Relative risk of dying before 65.", fig.height = 10, out.height = "18cm", cache = FALSE}
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1), lwd = 1.5)
midper <- c(1812.5, 1837.5, 1862.5, 1887.5, 1912.5, 1937.5, 1981, 1991, 2000.5, 2009.5)
refclw <- pdw[3, ]
plot(midper, pdw[1, ] / refclw, col = scol[1], type = "b", ylim = c(1/10, 5),
     xlab = "", xlim = c(1800, 2015), log = "y",
     ylab = "Relative risk", axes = FALSE, main = "Women, 40-64", lty = 1, pch = 1)

axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.8)
axis(2, las = 1)
axis(4, las = 1)
abline(h = 0)
box()
for (i in 2:5){
    lines(midper, pdw[i, ] / refclw, col = scol[i], type = "b", pch = i, lty = i)
}
legend("bottom", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 1.0, bty = "n")
refclm <- pdm[3, ]
plot(midper, pdm[1, ] / refclm, col = scol[1], type = "b", ylim = c(1/10, 5), xlab = "",
     ylab = "Relative risk", axes = FALSE, main = "Men, 40-64", lty = 1, pch = 1,
     xlim = c(1800, 2015), log = "y")
axis(1, at = midper, labels = labb, las = 2, cex.axis = 0.8)
axis(2, las = 1)
axis(4, las = 1)
abline(h = 0)
box()
for (i in 2:5){
    lines(midper, pdm[i, ] / refclm, col = scol[i], type = "b", lty = i, pch = i)
}
legend("bottom", legend = his, col = scol, lty = 1:5, pch = 1:5, cex = 1.0, bty = "n")
##woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|30)",
  ##              main = "Women 30-64", ylim = c(0, 1))##, exclude = 5:6)
##moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|30)",
  ##              main = "Men 30-64", ylim = c(0, 1))##, exclude = 5:6)
par(oldpar)
```

```{r cumswtab, message=FALSE, eval = TRUE}
fitt <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex + civst + urban, data = vb1)

# vb3064$social_class <- vb3064$hisclass
# vb3064$marital_status <- vb3064$civst
# vb3064$urban_or_rural <- vb3064$urban
# vb3064$age_group <- vb3064$age
# fitt <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
#             data = vb3064, family = poisson)
dr <- drop1(fitt, test = "Chisq")
# ##dr
```

```{r cumswtabout, message = FALSE, cache = FALSE, results='asis'}
##knitr::kable(dr, caption = "Significance tests, ages 40-64.", digits = c(0, 1, 2, 4))
print(xtable(dr,label = "tab:cumswtabout", caption = "ANOVA, ages 40--64."),
      type = "latex", comment = FALSE)
```

@@

We now turn to the elderly, those who in most cases have left the workforce.
We use the *expected remaining years of life at 65 up to age 100* (reported as mean age at death) as the summary measure, since the relative risk is a not so good measure when risks are high. First, looking at Table \@ref(tab:cumswtaboldout), it is obvious that the statistical significance regarding differences between groups over time is strong.

```{r cumswtabold}
# vb6599$marital_status <- vb6599$civst
# vb6599$social_class <- vb6599$hisclass
# vb6599$urban_or_rural <- vb6599$urban
# vb6599$age_group <- vb6599$age
# fittold <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
#             data = vb6599, family = poisson)

fittold <- coxph(Surv(enter, exit, event) ~ hisclass * period * sex + civst + urban, data = vb2)
drold <- drop1(fittold, test = "Chisq")
```

```{r cumswtaboldout, cap = "Table", cache = FALSE, results='asis'}
##knitr::kable(drold, caption = "Single term deletions, ages 65-99.", digits = c(0, 1, 2, 4))
##drold
print(xtable(drold,label = "tab:cumswtaboldout", caption = "ANOVA, ages 65--99."),
      type = "latex", comment = FALSE)
```

@@

We thank the Demographic Data Base, Umeå University, for giving us access to data and
Jonas Helgertz at the Department of Economic History, Lund University, for
providing us with scripts for creating Hisclass from Swedish census
data.
-->

# Appendix {-}

The exposure times in years for women and men are shown in Tables \@ref(tab:womenexp) and \@ref(tab:menexp), respectively.

```{r exposuretab, message = FALSE, cache = FALSE, results = 'asis'}
##expo <- aggregate(vb[, c("exposure")], by = vb[, c("hisclass", "period", "sex")], FUN = sum)
ex <- with(vb, tapply(exposure, list(period, hisclass, sex), sum))
print(xtable(ex[,,2], label = "tab:womenexp", caption = "Exposure time in years by social class and period, women.", digits = 0), 
      type = "latex", comment = FALSE)
```

```{r exposuretab2, message = FALSE, cache = FALSE, results = 'asis'}
print(xtable(ex[,,1], label = "tab:menexp", caption = "Exposure time in years by social class and period, men.", digits = 0), 
      type = "latex", comment = FALSE)
```

The number of deaths for women are shown in Table \@ref(tab:womendeaths), and for men in Table \@ref(tab:mendeaths).

```{r deathtab, message = FALSE, cache = FALSE, results = 'asis'}
##expo <- aggregate(vb[, c("exposure")], by = vb[, c("hisclass", "period", "sex")], FUN = sum)
deaths <- with(vb, tapply(event, list(period, hisclass, sex), sum))
print(xtable(deaths[,,2], label = "tab:womendeaths", caption = "Number of deaths by social class and period, women.", digits = 0), 
      type = "latex", comment = FALSE)
```

```{r deathtab2, message = FALSE, cache = FALSE, results = 'asis'}
print(xtable(deaths[,,1], label = "tab:mendeaths", caption = "Number of deaths by social class and period, men.", digits = 0), 
      type = "latex", comment = FALSE)
```
